---
title: "Deployment"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

This guide shows how to deploy MOSAIC on remote virtual machines or compute clusters for production use and large-scale simulations. MOSAIC provides ready-to-use bash scripts that handle all installation steps automatically.

The deployment scripts are:
- **Non-interactive** - Run without user input, perfect for automation
- **SSH-compatible** - Execute remotely via `ssh user@host 'bash -s' < script.sh`
- **Cloud-ready** - Use with cloud-init or VM provisioning tools
- **Verified** - Include automatic dependency checking

All scripts are available in the `vm/` directory of the MOSAIC-pkg repository.

---

## Complete VM Setup

The `vm/setup_mosaic.sh` script installs R, system dependencies, MOSAIC package, and Python components from scratch on Ubuntu/Debian systems.

**Execute via SSH:**
```{sh vm-setup, eval=FALSE}
ssh user@host 'bash -s' < vm/setup_mosaic.sh
```

**Or run directly on VM:**
```{sh vm-direct, eval=FALSE}
bash vm/setup_mosaic.sh
```

**Script contents:**

```{bash, code=readLines('../../vm/setup_mosaic.sh'), eval=FALSE}
```

---

## Minimal VM Setup

If R >= 4.1.1 is already installed on your VM, use `vm/setup_mosaic_minimal.sh`:

**Execute via SSH:**
```{sh vm-minimal, eval=FALSE}
ssh user@host 'bash -s' < vm/setup_mosaic_minimal.sh
```

**What it does:**

This script is identical to the complete setup but skips R installation (step 2/6). It installs:

- System libraries (GDAL, PROJ, GEOS, UDUNITS, Python)
- MOSAIC R package from GitHub
- Python dependencies (laser-cholera)
- Verifies installation with `check_dependencies()`

---

## Parallel Execution

For running multiple simulations in parallel on a cluster:

```{r vm-parallel}
# Set number of cores
library(MOSAIC)
options(mc.cores = parallel::detectCores() - 1)

# Run parallel simulations
ctrl <- mosaic_control(
  parallel = TRUE,
  n_cores = parallel::detectCores() - 1
)

results <- run_LASER(
  config = config_default,
  control = ctrl,
  n_sim = 100,
  seed = 123
)
```

---

## Troubleshooting

**Python issues:**
```{r troubleshoot}
# Check Python configuration
reticulate::py_config()

# Reset if needed
remove_MOSAIC_python_env()
install_dependencies()
```

---

## Next Steps

After deployment, see the **"Running MOSAIC"** vignette to learn model execution.
