---
title: "Deployment"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Overview

Deploy MOSAIC on remote virtual machines or compute clusters using transportable bash scripts. All scripts are non-interactive and can be executed via SSH or cloud-init.

Setup scripts are available in the `vm/` directory of the MOSAIC-pkg repository.

---

## Complete VM Setup

The `vm/setup_mosaic.sh` script installs R, system dependencies, MOSAIC package, and Python components from scratch on Ubuntu/Debian systems.

**Execute via SSH:**
```{sh vm-setup}
ssh user@host 'bash -s' < vm/setup_mosaic.sh
```

**Or run directly on VM:**
```{sh vm-direct}
bash vm/setup_mosaic.sh
```

**What it installs:**
- System updates
- R (>= 4.1.1) from CRAN
- Geospatial libraries (GDAL, PROJ, GEOS, UDUNITS)
- Additional dev libraries (libcurl, libssl, libxml2, etc.)
- Python 3.9+
- MOSAIC R package from GitHub
- Python dependencies (laser-cholera)

---

## Minimal VM Setup

If R >= 4.1.1 is already installed on your VM, use `vm/setup_mosaic_minimal.sh`:

**Execute via SSH:**
```{sh vm-minimal}
ssh user@host 'bash -s' < vm/setup_mosaic_minimal.sh
```

**What it installs:**
- System libraries only (GDAL, PROJ, GEOS, UDUNITS, Python)
- MOSAIC R package
- Python dependencies

---

## Parallel Execution

For running multiple simulations in parallel on a cluster:

```{r vm-parallel}
# Set number of cores
library(MOSAIC)
options(mc.cores = parallel::detectCores() - 1)

# Run parallel simulations
ctrl <- mosaic_control(
  parallel = TRUE,
  n_cores = parallel::detectCores() - 1
)

results <- run_LASER(
  config = config_default,
  control = ctrl,
  n_sim = 100,
  seed = 123
)
```

---

## Troubleshooting

**Python issues:**
```{r troubleshoot}
# Check Python configuration
reticulate::py_config()

# Reset if needed
remove_MOSAIC_python_env()
install_dependencies()
```

---

## Next Steps

After deployment, see the **"Running MOSAIC"** vignette to learn model execution.
